Type-safe client-server communication for functional web applications

### 2014-02-28 paper.tex
Finish explaining Client monad.

The \lstinline!export! function takes an arbitrary function, provided that all
its arguments as well as its return value are serializable through the aptly
named \lstinline!Serialize! type class, and returns a typed ``handle'' of
sorts. In this example, the type of \lstinline!greetings! is
\lstinline!Export (String -> Server ())!, indicating that the handle identifies
an exported function with a single \lstinline!String! argument and no return
value; exported functions all live in the \lstinline!Server! monad.

@@

\begin{code}
srvSend :: Useless RcptList -> String -> Server ()
srvSend uRcpts message = do
    rcpts <- mkUseful uRcpts
    liftIO $ do
      recipients <- CC.withMVar rcpts return
      mapM_ (CC.forkIO . deliver message) recipients
  where
    deliver msg (_, rcptMVar) = CC.putMVar rcptMVar msg
\end{code}

@@

Aside from the \lstinline!runClient! function, which is only used to launch the
client program, This monad has two primary operations.

@@

@@

@@

We have presented a programming model which improves on the current state of
the art in client-server web application development. In particular, our
solution combines type safe communication between the client and the server
with purely functional semantics, clear demarcations as to when data is
transmitted and where a particular piece of code is executed, and the ability to
effortlessly share code between the client and the server.

Our model is based on a simple blocking concurrency model rather than explicit
continuations which is well suited for use with a GUI programming style based on
communicating processes with local state, and requires essentially no
modification of existing tools or compilers.

@@

\bibitem[Svenningsson, Svensson (2013)]{bjornbenny}
J. Svenningsson, J. Svensson: Simple and Compositional Reification of Monadic Embedded Languages, 2013

>>>>

The \lstinline!export! function takes an arbitrary function, provided that all
its arguments as well as its return value are serializable through the aptly
named \lstinline!Serialize! type class, and returns a typed ``handle'' of
sorts. In this example, the type of \lstinline!greetings! is
\linebreak
\lstinline!Export (String -> Server ())!, indicating that the handle identifies
an exported function with a single \lstinline!String! argument and no return
value; exported functions all live in the \lstinline!Server! monad.

@@

\begin{code}
srvSend :: Useless RcptList -> String -> Server ()
srvSend uRcpts message = do
    rcpts <- mkUseful uRcpts
    liftIO $ do
      recipients <- CC.withMVar rcpts return
      mapM_ (CC.forkIO . deliver message) recipients
  where
    deliver msg (_,rcptMVar) = CC.putMVar rcptMVar msg
\end{code}

@@

What makes \lstinline!App! interesting is that computations in this monad are
executed on both the client and the server; once on server startup, and once
in the startup phase of each client. Its operations behave slightly differently
depending on whether they are executed on the client or on the server.
Execution is deterministic, ensuring that the same sequence of
\lstinline!CallID!s are generated during every run, both on the server and on
all clients. This is necessary to ensure that any particular call identifier
always refers to the same server-side function on all clients.

After all common code has been executed, the program flow diverges between the
client and the server; client-side, \lstinline!runClient! launches the
application's \lstinline!Client! computation whereas on the server, this
computation is discarded, and the server instead goes into an event loop,
waiting for calls from the client.

Aside from \lstinline!runClient!, this monad has two primary operations.

@@

After the \lstinline!App! computation finishes, the identifier-function mapping
accumulated in its state is handed over to the server's event loop, where it is
used to dispatch the proper functions for incoming calls from the client.

@@

\begin{code}
type Nonce = Int
type Client = StateT Nonce Conc
\end{code}

Aside from the added concurrency capabilities, the \lstinline!Client! monad
only has a single particularly interesting operation: \lstinline!onServer!.

\begin{code}
onServer :: Serialize a
         => Export (Server a) -> Client a
onServer (Export identifier args) = do
  nonce <- get
  put (nonce + 1)
  mv <- createResultMVar nonce
  webSocketSend $ serialize (nonce, identifier, reverse args)  
  takeMVar mv
\end{code}

A nonce is used to pair up return values with the calls that produced them.
After a call is dispatched, \lstinline!onServer! blocks, waiting for its
\emph{result variable} to be filled with the result of the call. Filling this
variable is the responsibility of the \emph{receive callback}, which is
executed every time a message arrives from the server.

\begin{code}
onMessage :: JSON -> Client ()
onMessage msg = do
  let (nonce, result) = deserialize msg
  mv <- findResultMVar nonce
  putMVar mv result
\end{code}

The corresponding server-side event loop, which handles incoming calls from the
client, is similarly straightforward.

\begin{code}
onEvent :: [(CallID, Method)] -> JSON -> IO ()
onEvent mapping msg = do
  let (nonce, identifier, args) = deserialize msg
      f = find identifier mapping
  result <- f args
  webSocketSend $ serialize (nonce, result)
\end{code}

The function corresponding to the RPC call's identifier is looked up in the
identifier-function mapping generated by the \lstinline!App! computation and
applied to the list of arguments; the return value is bundled up with the
call's nonce and sent back to the client.

@@

We have presented a programming model which improves on the current state of
the art in client-server web application development. In particular, our
solution combines type safe communication between the client and the server
with functional semantics, clear demarcations as to when data is transmitted
and where a particular piece of code is executed, and the ability to
effortlessly share code between the client and the server.

Our model is based on a simple blocking concurrency model rather than explicit
continuations which is well suited for use with a GUI programming style based on
self-contained processes with local state, and requires essentially no
modification of existing tools or compilers.

@@
